Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge

### 2019-05-31 man/ecosystems/ms_ecosystems.Rmd
update

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal tree-line in the sampled slope. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].

### Dendroecological analyses 
Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally done using COFECHA software [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package. Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].

@@

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018]. The Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014], was also used to characterize the drought conditions for the period 1961-2014. SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from the Drought indices dataset (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].

Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015]. The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging, neighbor mortality) in the past. By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].

Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

@@

In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@MartinMontanes2015; @RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].

## Resilience

To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and before the disturbance (drought) event:

@@

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017].

@@

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Figure 3).

## Radial growth trends and growth disturbances

>>>>

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.

Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally done using COFECHA software [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].

@@

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Frias2018]. The Standardized Precipitation-Evapotranspiration Index (SPEI)  [@VicenteSerrano2010] was used to characterize the drought conditions for the period 1961-2014. SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from a high resolution database of drought indices for Spain  [@VicenteSerrano2017].

Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using `treeclim` package [@Zang2015].

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging, neighbor mortality) in the past. By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997]. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

@@

In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].

## Assessing ecosystem and tree individual resilience to drought

To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and before the disturbance (drought) event:

@@

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` and `rcompanion`  packages. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).

## Analysis of radial growth trends and disturbances
